Overrelaxed Sinkhorn--Knopp Algorithm for Regularized Optimal Transport

### 2017-11-25 PosterNIPS2017.tex
Poster

		The algorithm we obtain converges faster than Sinkhorn--Knopp, while remaining parallelizable, and having almost the same cost per iteration.
	\end{minipage}
	\begin{minipage}{0.45\textwidth}
		\centering
		\begin{tcolorbox}[width=0.99\textwidth,/tcb/size=small, /tcb/top=2.0mm]
			\centering
			\begin{tabular}{c c c c}
				\includegraphics[height=3cm]{1d_interp_mu} &
				\includegraphics[height=3cm]{1d_interp_0_2} &
				\includegraphics[height=3cm]{1d_interp_2} &
				\includegraphics[height=3cm]{1d_interp_18} \\
				&
				$\epsilon = 5$ &
				$\epsilon = 0.5$ &
				$\epsilon = 0.02$
			\end{tabular}
			\captionof{figure}{Regularized optimal transport maps for decreasing regularization parameter $\epsilon$.} \label{fig:1d_ot}
		\end{tcolorbox}
	\end{minipage}
}

@@

\headerbox{Overrelaxation}{name=overrelaxation,column=2}{
\begin{minipage}{0.49\textwidth}
	\begin{tcolorbox}[size=small, top=2mm]
		\centering
		\begin{tabular}{c c}
		\input{schema_b} &
		\input{schema_c} \\
		SK &
		Overrelaxed SK
		\end{tabular}
	\captionof{figure}{For small regularization $\epsilon$, the SK algorithm converges slowly, whereas the overrelaxed SK algorithm converges with a better rate}
	\end{tcolorbox}
\end{minipage}
\begin{minipage}{0.49\textwidth}
	\vspace{0.6mm}
	Although the SK algorithm is much more efficient than linear solvers in a high-dimensional setting, it becomes extremely slow when the regularization $\epsilon$ is small.
	To counteract this limitation, we use the principle of \emph{overrelaxation}.

	We define the overrelaxed projections of parameter $\theta \in \IR$:
	\begin{align*}\label{or_scaling}
	P_{\Ccal_1}^\theta(\gamma) &= \diag(a)^\theta \gamma\\
	P_{\Ccal_2}^\theta(\gamma) &= \gamma \diag(b)^\theta \nonumber .
	\end{align*}
	In these equations, the exponentiation is applied element-wise.

	In the euclidean setting, which also corresponds to the linearization of the operators around the solution, iterative overrelaxed projections can be shown to converge for any $\theta \in (0,2)$.
\end{minipage}
}

\headerbox{Experiments}{name=expe,below=overrelaxation,column=2}{
	\begin{minipage}{\textwidth}
		We compare our algorithm to SK algorithm on two different optimal transport settings.
		\begin{itemize}
			\item In setting (a) we consider the domain $[0,1]$ discretized into $100$ samples and the squared Euclidean transport cost on this domain.
			The marginals are densities made of the sum of a base plateau of height $0.1$ and another plateau of height and boundaries chosen uniformly in $[0,1]$, subsequently normalized.
			\item In setting (b) the cost is a $100\times 100$ random matrix with entries uniform in $[0,1]$ and the marginals are uniform.
		\end{itemize}
		We plot the relative speed of our algorithm compared to Sinkhorn, for various values of the regularization parameter $\epsilon$. In both experimental settings, our algorithm wins by a factor above $20$ for small values of $\epsilon$.
	\end{minipage}
	\begin{minipage}{\textwidth}
		\begin{tcolorbox}[size=small]
			\begin{minipage}[b]{.5\linewidth}
				\centering
				\includegraphics[scale=0.4]{images/speedratio_image}\\
				{\centering\small (a) Quadratic cost, random marginals}\label{fig:compare_a}
			\end{minipage}%
			\begin{minipage}[b]{.5\linewidth}
				\centering
				\includegraphics[scale=0.4]{images/speedratio_ML}\\
				{\centering\small (b) Random cost, uniform marginals}\label{fig:compare_b}
			\end{minipage}
		\end{tcolorbox}
	\end{minipage}
}

\headerbox{References}{name=references,below=expe,column=2}{
	\small
	\begin{tabular}{r p{0.67\textwidth}}
		{[Cuturi, 2013]} & M. Cuturi. Sinkhorn distances: Lightspeed computation of optimal transport. {\it Advances in Neural Information Processing Systems 26}, 2013.\\
		{[Benamou et al., 2015]} & Benamou, J.-D., Carlier, G., Cuturi, M., Nenna, L., and Peyr\'e, G. (2015).
		{\it Iterative Bregman projections for regularized transportation problems.} SIAM Journal on Scientific
		Computing, 37(2):A1111–A1138.\\
		{[Sinkhorn, 1964]} & Sinkhorn, R. (1964). {\it A relationship between arbitrary positive matrices and doubly stochastic matrices.} The annals of mathematical statistics, 35(2):876–879.
	\end{tabular}
}

@@

\headerbox{Sinkhorn-Knopp Algorithm}{name=Sinkhorn,column=0,below=intro,bottomaligned=references}{
	We aim at numerically solving the entropic regularization of optimal transport:
	\begin{equation} \label{eq:problem}
	\gamma^* = \argmin_{\gamma \in \Ccal_1 \cap \Ccal_2}
	\scal{c}{\gamma} + \epsilon \KL(\gamma,\One).
	\end{equation}
	In this equation, $\Ccal_1$ and $\Ccal_2$ are the marginal constraint sets corresponding to two discrete measures $\mu^1$ and $\mu^2$,
	and where $\KL$ is the Kullback--Leibler divergence.
	\begin{gather*}
	\Ccal_1 = \left\{ \gamma \mid A_1 \gamma = \mu^1 \right\},
	\quad\quad\quad
	\Ccal_2 = \left\{ \gamma \mid A_2 \gamma = \mu^2 \right\}.\\
	\KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}.
	\end{gather*}
	\begin{minipage}{0.6\textwidth}
		As problem (\ref{eq:problem}) is strictly convex, it has exactly one solution, that can be written equivalently as the \emph{Bregman projection} of $\gamma^0 = e^{-c/\epsilon}$ onto a convex set:
		\begin{gather*}
		P_{\Ccal}(\xi) := \argmin_{\gamma \in \Ccal} \KL(\gamma,\xi),\\
		\text{(\ref{eq:problem})} \quad \Longleftrightarrow \quad \gamma^* = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0).
		\end{gather*}
		Whereas projecting onto $\Ccal_1 \cap \Ccal_2$ is challenging, projecting onto $\Ccal_1$ or $\Ccal_2$ individually is simply performed by scaling the rows or columns of the matrix so that the corresponding marginal sum is correct.
		Operator $\oslash$ denotes element-wise division:
		\begin{align*}\label{scaling}
		P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &\text{with}\quad
		a &=  {\mu^1}\oslash{A_1 \gamma} \\
		P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &\text{with}\quad
		b &= {\mu^2}\oslash{A_2 \gamma}\nonumber
		\end{align*}
	\end{minipage}
	\begin{minipage}{0.39\textwidth}
		\begin{tcolorbox}[/tcb/size=small, /tcb/top=2.0mm]
			\centering
			\input{schema_a}
			\captionof{figure}{The Sinkhorn--Knopp algorithm iterates Bregman projections on the two constraint sets.}
		\end{tcolorbox}
	\end{minipage}

@@

\headerbox{Lyapunov Function}{name=lyapunov,column=1,below=intro,bottomaligned=Sinkhorn}{
	There is no general proof of global convergence for iterative overrelaxed Bregman projections. In order to ensure convergence, we use a Lyapunov function.
	\begin{tcolorbox}[size=small]
    	Choose $F(\gamma)$ continuous, coercive function whose unique minimizer is $\gamma^*$.\\
    	Choose $\theta_1(\gamma)$, $\theta_2(\gamma)$ continuous functions, such that
    	\begin{equation*}\label{eq:cond_theta_k}
    	\forall k \in \{1,2\}, \, \forall \gamma \notin \Ccal_k,\quad\quad
    	F(P_{\Ccal_k}^{\theta_k(\gamma)}(\gamma)) < F(\gamma).
    	\end{equation*}
    	Alternately iterate $P_{\Ccal_1}^{\theta_1}$ and $P_{\Ccal_2}^{\theta_2}$, starting from $\gamma^0 = e^{-c/\epsilon}$.\\
    	This process converges to $\gamma^*$.
    \end{tcolorbox}
    Our choice of Lyapunov function is: \[F(\gamma) := \KL(\gamma^*,\gamma)\]
    It should be noted that this function decreases for regular Bregman projections.

    \begin{minipage}{0.35\textwidth}
    	\begin{tcolorbox}[size=small]
    	\centering
    	\includegraphics[width=\textwidth]{images/cvgce_zone_2.png}
    	\captionof{figure}{
    		Function $\varphi_\theta(x)$ is positive above the red line, negative below.}
    	\end{tcolorbox}
    \end{minipage}
    \begin{minipage}{0.65\textwidth}
    	The difference between two evaluations of the function can be computed directly thanks to the formula:
    	\begin{equation*} \label{eq:kl_diff_scal}
    	F(\gamma) - F(P^\theta_{\Ccal_k}(\gamma)) = 
    	\scal{\mu^k}{\varphi_\theta \left((A_k \gamma) \oslash \mu^k \right)}.
    	\end{equation*}
    	We want this difference to be positive. A sufficient condition is that every coordinate of $\varphi_\theta \left((A_k \gamma) \oslash \mu^k \right)$ be positive. Functions $\theta_1, \theta_2$ are crafted for this purpose.
    	\begin{gather*}
    	\Theta^*(u) := \max \left\{\theta \mid \varphi_\theta(u) \ge 0 \right\}
    	\\
    	\Theta(u) := clip_{[1,\theta_0]}(\Theta^*(\min u)-\delta)\\
    	k \in \{1,2\}, \quad \theta_k (\gamma) := \Theta((A_k \gamma) \oslash \mu^k)
    	\end{gather*}
    \end{minipage}
    This adaptive choice allows to choose the overrelaxation parameter $\theta_0$ that is used asymptotically. A linear study around the solution shows that overrelaxation modifies the eigenvalues in a predictable way: in particular, if the linear convergence rate of the Sinkhorn--Knopp algorithm is $1-\eta$, then the optimal asymptotic overrelaxation parameter is $\theta_0 = \frac{2}{1+\sqrt{\eta}}$.
}

>>>>

		The algorithm we obtain converges between 2 and 20 times faster than Sinkhorn--Knopp, while remaining parallelizable, and having almost the same cost per iteration.
	\end{minipage}
	\begin{minipage}{0.45\textwidth}
		\centering
		\begin{tcolorbox}[width=0.99\textwidth,/tcb/size=small, /tcb/top=2.0mm]
			\centering
			\begin{tabular}{c c c c}
				\includegraphics[height=3cm]{1d_interp_mu} &
				\includegraphics[height=3cm]{1d_interp_0_2} &
				\includegraphics[height=3cm]{1d_interp_2} &
				\includegraphics[height=3cm]{1d_interp_18} \\
				&
				$\epsilon = 5$ &
				$\epsilon = 0.5$ &
				$\epsilon = 0.02$
			\end{tabular}
			\captionof{figure}{Regularized optimal transport maps for decreasing regularization parameter $\epsilon$.} \label{fig:1d_ot}
		\end{tcolorbox}
	\end{minipage}
}

@@

\headerbox{Overrelaxation}{name=overrelaxation,column=1,below=intro}{
\begin{minipage}{0.49\textwidth}
	\begin{tcolorbox}[size=small, top=2mm]
		\centering
		\begin{tabular}{c c}
		\input{schema_b} &
		\input{schema_c} \\
		SK &
		Overrelaxed SK
		\end{tabular}
	\captionof{figure}{For small regularization $\epsilon$, the SK algorithm converges slowly, whereas the overrelaxed SK algorithm converges with a better rate}
	\end{tcolorbox}
\end{minipage}
\begin{minipage}{0.49\textwidth}
	\vspace{0.6mm}
	Although the SK algorithm is much more efficient than linear solvers in a high-dimensional setting, it becomes extremely slow when the regularization $\epsilon$ is small.
	To counteract this limitation, we use the principle of \emph{overrelaxation}.

	We define the overrelaxed projections of parameter $\theta \in \IR$:
	\begin{align*}\label{or_scaling}
	P_{\Ccal_1}^\theta(\gamma) &= \diag(a)^\theta \gamma\\
	P_{\Ccal_2}^\theta(\gamma) &= \gamma \diag(b)^\theta \nonumber .
	\end{align*}
	In these equations, the exponentiation is applied element-wise.
\end{minipage}
As in the Sinkhorn--Knopp algorithm, we iterate these operators alternately.
If we consider the linearization of the operators around the solution, iterative overrelaxed projections can be shown to converge for any $\theta \in (0,2)$.
In the global, non-euclidean setting, we use an additional criterion based on a Lyapunov function.
}

@@

\headerbox{Sinkhorn-Knopp Algorithm}{name=Sinkhorn,column=0,below=intro}{
	We aim at numerically solving the entropic regularization of optimal transport:
	\begin{equation} \label{eq:problem}
	\gamma^* = \argmin_{\gamma \in \Ccal_1 \cap \Ccal_2}
	\scal{c}{\gamma} + \epsilon \KL(\gamma,\One).
	\end{equation}
	In this equation, $\Ccal_1$ and $\Ccal_2$ are the marginal constraint sets corresponding to two discrete measures $\mu^1$ and $\mu^2$,
	and where $\KL$ is the Kullback--Leibler divergence.
	\begin{gather*}
	\Ccal_1 = \left\{ \gamma \mid A_1 \gamma = \mu^1 \right\},
	\quad\quad\quad
	\Ccal_2 = \left\{ \gamma \mid A_2 \gamma = \mu^2 \right\}.\\
	\KL(\gamma,\xi) = \sum_{i,j} \gamma_{i,j} \left( \log \left( \frac{\gamma_{i,j}}{\xi_{i,j}} \right) -1  \right) + \sum_{i,j} \xi_{i,j}.
	\end{gather*}
	\begin{minipage}{0.55\textwidth}
		As problem (\ref{eq:problem}) is strictly convex, it has exactly one solution, that can be written equivalently as the \emph{Bregman projection} of $\gamma^0 = e^{-c/\epsilon}$ onto a convex set:
		\begin{gather*}
		P_{\Ccal}(\xi) := \argmin_{\gamma \in \Ccal} \KL(\gamma,\xi),\\
		\text{(\ref{eq:problem})} \quad \Longleftrightarrow \quad \gamma^* = P_{\Ccal_1 \cap \Ccal_2}(\gamma^0).
		\end{gather*}
		Whereas projecting onto $\Ccal_1 \cap \Ccal_2$ is challenging, projecting onto $\Ccal_1$ or $\Ccal_2$ individually is simply performed by scaling the rows or columns of the matrix so that the corresponding marginal sum is correct.
		Operator $\oslash$ denotes element-wise division:
		\begin{align*}\label{scaling}
		P_{\Ccal_1}(\gamma) &= \diag(a) \gamma &\text{with}\quad
		a &=  {\mu^1}\oslash{A_1 \gamma} \\
		P_{\Ccal_2}(\gamma) &= \gamma \diag(b) &\text{with}\quad
		b &= {\mu^2}\oslash{A_2 \gamma}\nonumber
		\end{align*}
	\end{minipage}
	\begin{minipage}{0.44\textwidth}
		\begin{tcolorbox}[/tcb/size=small, /tcb/top=2.0mm]
			\centering
			\input{schema_a}
			\captionof{figure}{The Sinkhorn--Knopp algorithm iterates Bregman projections on the two constraint sets.}
		\end{tcolorbox}
	\end{minipage}

@@

\headerbox{Lyapunov Function}{name=lyapunov,column=2}{
	Although the overrelaxation of iterative Bregman projections as presented here is already in use in the context of numerical calculations of optimal transport, there is no proof that this algorithm converges to the solution. In order to ensure this result, we use a Lyapunov function, and slightly modify the algorithm.
	\begin{tcolorbox}[size=small]
    	Choose $F(\gamma)$ continuous, coercive function whose unique minimizer is $\gamma^*$.\\
    	Choose $\theta_1(\gamma)$, $\theta_2(\gamma)$ continuous functions, such that
    	\begin{equation*}\label{eq:cond_theta_k}
    	\forall k \in \{1,2\}, \, \forall \gamma \notin \Ccal_k,\quad\quad
    	F(P_{\Ccal_k}^{\theta_k(\gamma)}(\gamma)) < F(\gamma).
    	\end{equation*}
    	Alternately iterate $P_{\Ccal_1}^{\theta_1}$ and $P_{\Ccal_2}^{\theta_2}$, starting from $\gamma^0 = e^{-c/\epsilon}$.\\
    	This process converges to $\gamma^*$.
    \end{tcolorbox}
    Our choice of Lyapunov function is: \[F(\gamma) := \KL(\gamma^*,\gamma)\]
    It should be noted that this function decreases for regular Bregman projections.

    The difference between two evaluations of the function can be computed directly thanks to the formula:

    \begin{minipage}{0.65\textwidth}
    	\begin{equation*} \label{eq:kl_diff_scal}
    	F(\gamma) - F(P^\theta_{\Ccal_k}(\gamma)) = 
    	\scal{\mu^k}{\varphi_\theta \left((A_k \gamma) \oslash \mu^k \right)}.
    	\end{equation*}
    	We want this difference to be positive. A sufficient condition is that every coordinate of $\varphi_\theta \left((A_k \gamma) \oslash \mu^k \right)$ be positive. Functions $\theta_1, \theta_2$ are crafted for this purpose.
    	\begin{gather*}
    	\Theta^*(u) := \max \left\{\theta \mid \varphi_\theta(u) \ge 0 \right\}
    	\\
    	\Theta(u) := clip_{[1,\theta_0]}(\Theta^*(\min u)-\delta)\\
    	k \in \{1,2\}, \quad \theta_k (\gamma) := \Theta((A_k \gamma) \oslash \mu^k)
    	\end{gather*}
    \end{minipage}
	\begin{minipage}{0.34\textwidth}
		\begin{tcolorbox}[size=small]
			\centering
			\includegraphics[width=\textwidth]{images/cvgce_zone_2.png}
			\captionof{figure}{
				Function $\varphi_\theta(x)$ is positive above the red line, negative below.}
		\end{tcolorbox}
	\end{minipage}
This adaptive choice allows to choose the overrelaxation parameter $\theta_0$ that is used asymptotically. A linear study around the solution shows that overrelaxation modifies the eigenvalues in a predictable way: in particular, if the linear convergence rate of the Sinkhorn--Knopp algorithm is $1-\eta$, then the optimal asymptotic overrelaxation parameter is $\theta_0 = \frac{2}{1+\sqrt{\eta}}$.
}

\headerbox{Experiments}{name=expe,below=lyapunov,column=2}{
	\begin{minipage}{\textwidth}
		We compare our algorithm to SK algorithm on two different settings.
		\begin{itemize}
			\item In setting (a) we consider the domain $[0,1]$ discretized into $100$ samples and the squared Euclidean cost.
			The marginals are randomly generated, piecewise constant on $[0,1]$.
			\item In setting (b) the cost is a $100\times 100$ random matrix with entries uniform in $[0,1]$ and the marginals are uniform.
		\end{itemize}
		We plot the relative speed of our algorithm compared to Sinkhorn, for various values of the regularization parameter $\epsilon$. In both experimental settings, our algorithm wins by a factor above $20$ for small values of $\epsilon$.
	\end{minipage}
	\begin{minipage}{\textwidth}
		\begin{tcolorbox}[size=small]
			\begin{minipage}[b]{.5\linewidth}
				\centering
				\includegraphics[scale=0.35]{images/speedratio_image}\\
				{\centering\small (a) Quadratic cost, random marginals}\label{fig:compare_a}
			\end{minipage}%
			\begin{minipage}[b]{.5\linewidth}
				\centering
				\includegraphics[scale=0.35]{images/speedratio_ML}\\
				{\centering\small (b) Random cost, uniform marginals}\label{fig:compare_b}
			\end{minipage}
		\end{tcolorbox}
	\end{minipage}
}

\headerbox{References}{name=references,below=overrelaxation,column=1}{
	\small
	\begin{tabular}{r p{0.67\textwidth}}
		{[Cuturi, 2013]} & M. Cuturi. Sinkhorn distances: Lightspeed computation of optimal transport. {\it Advances in Neural Information Processing Systems 26}, 2013.\\
		{[Benamou et al., 2015]} & Benamou, J.-D., Carlier, G., Cuturi, M., Nenna, L., and Peyr\'e, G. (2015).
		{\it Iterative Bregman projections for regularized transportation problems.} SIAM Journal on Scientific
		Computing, 37(2):A1111–A1138.\\
		{[Sinkhorn, 1964]} & Sinkhorn, R. (1964). {\it A relationship between arbitrary positive matrices and doubly stochastic matrices.} The annals of mathematical statistics, 35(2):876–879.
	\end{tabular}
}
